---
title: "raerdata: a collection of datasets and databases for use with the raer package" 
author: 
  - name: Kent Riemondy
    affiliation: University of Colorado School of Medicine
date: '`r Sys.Date()`'
output:
  BiocStyle::html_document
package: raerdata 
bibliography: ref.bib 
vignette: >
  %\VignetteIndexEntry{raerdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

The `raerdata` package contains datasets and databases used to illustrate functionality
in the `raer` package.

```{r}
library(raerdata)
```

The package contains the following resources.

## RNA editing Atlases 

Atlases of known RNA editing sites formatted into `GRanges` objects. 

### REDIportal 

The `REDIportal` is a collection of RNA editing sites identified from multiple studies
in multiple species (@Picardi2017-gn). The human (`hg38`) and mouse (`mm10`) collections are provided in 
GRanges objects, in either coordinate only format, or with additional metadata. 

```{r}
rediportal_coords_hg38()
```

### CDS recoding sites 

Human `CDS` recoding RNA editing sites identified by @Gabay2022-gw were formatted into
`GRanges` objects. These sites were also lifted over to the mouse genome (`mm10`).

```r
gabay_sites_hg38()
```

## Datasets

### Whole genome and RNA sequencing data from NA12877 cell line

WGS and RNA-seq BAM and associated files generated from a subset of chromosome 4.
Paths to files and data objects are returned in a list. 

```{r}
NA12877()
```

### GSE99249: RNA-Seq of interferon Beta treatment of ADAR1KO cell line

RNA-seq BAM files from ADAR1KO and Wild-Type HEK293 cells and associated files from chromosome 18 (@Chung2018-gh)

```{r}
GSE99249()
```


### 10x Genomics 10k PBMC scRNA-seq

10x genomics BAM file and associated files from chromosome 16 of human PBMC scRNA-seq library.

```{r}
pbmc_10x()
```

## ExperimentHub access

Alternatively individual files can be accessed from the ExperimentHub directly

```{r}
library(ExperimentHub)
eh <- ExperimentHub()
query(eh, "raerdata")
```

<details style="margin-bottom:10px;">
<summary>
    Session info
</summary>

```{r}
sessionInfo()
```

</details>


